---
title: "BCGcalc-NEWS"
author: "Erik.Leppo@tetratech.com"
date: "`r Sys.time()`"
output: github_document
  #md_document:
   # variant: markdown_github
---

<!-- NEWS.md is generated from NEWS.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "NEWS-"
)
```

```{r, echo = FALSE}
cat(paste0("Last Update: ",Sys.time()))
```

# Future Possibile Features (Wish List)

* ~~Add excluded taxa flag to samples based on levels of taxonomy.  Will not be completely accurate.  Will have many false positives and false negatives.  That is, if there are only organisms at the higher level in a sample it will fail.  And if there are no organisms of the target group it will pass.~~
    + ~~Pass, Fail on each. And overall flag~~
    + ~~Chiro (tribe and family)~~
    + ~~Oligo~~
    + ~~Trombidiformes~~
    
* Hi/Lo models.
    + ~~parameter to run "correct" model or all models for each site.~~
    + Have a field for site type.  So can override gradient.
    
* ~~Report - QC, use DataExplorer.~~

* ~~Table of metrics for each model with DataExplorer report.~~

* Include a master taxa list.
    + Match data input and add information (phylo and auteco)
    + Ability to report back unmatched
    
* Generate SiteType from data (gradient for PacNW)

* Update For loop in BCG.Level.Assignment to apply function (similar to BCG.Level.Membership).

* Any issues as reported on GitHub.  
    + https://github.com/leppott/BCGcalc/issues

# Version History

## Changes in Version 1.2.2.9003 (2018-09-04)

* metric.values.  Issue #16
    + Added QC worksheet to extdata\Data_BCG_PacNW.xlsx

## Changes in Version 1.2.2.9002 (2018-08-28)

* metric.values.  Issue #13
    + Add required fields (columns) to help file.
    + Add QC check in function.  Prompt user to continue or stop.

## Changes in Version 1.2.2.9001 (2018-08-15)

* metric.values
    + Added metric Percent Baetis tricaudatus complex + Simuliidae individuals (pi_SimBtri).  Issue #16.

## Changes in Version 1.2.2 (2018-06-14)

* New release version.

## Changes in Version 1.2.1.9001 (2018-06-14)

* qc.checks.R
    + Modification to use metric names as lower case.  Issue #12.
    + Example table; useNA="ifany".

## Changes in Version 1.2.1 (2018-06-13)

* New release version.

## Changes in Version 1.2.0.9001 (2018-06-13)

* BCG.Level.Assignment.R
    + Membership QC (sum===1) add "round" 8 to avoid too many samples being flagged as "FAIL".
    + Example changes:
    + Changed terminology from PASS/FAIL to NA/flag in flag fields.
    + Changed "NumFlagFAIL" to "NumFlags".
    + Reordered columns so "NumFlags" is before the flag fields.

## Changes in Version 1.2.0 (2018-06-13)

* Release version for Pacific Northwest BCG workgroup with updated model.

## Changes in Version 1.1.0.9010 (2018-06-13)

* BCG.Level.Assignment.R
    + Example code to match up with qc.checks output.
    + Includes export to file.
* qc.checks.R
    + Remove example modfying column names.

## Changes in Version 1.1.0.9009 (2018-06-13)

* BCG.Level.Membership.R
    + Added "round" to 8 places when determining level membership.
    + Floating point error resulted in the occassional value of 1.1E-16 when the previous level was "1".
    + This resulted in the sample getting a split level assignment, e.g., 3:4 when it was 100% 3.

## Changes in Version 1.1.0.9008 (2018-06-13)

* metric.values.R
    + Add additional metrics
    + nt; BCG_att; 1i, 1m, 3
    + pt; BCG_att; 1i, 1m, 3
    + pi; BCG_att; 1i, 1m, 2, 3
    + thermal indicator metrics
    + modify grepl to == because COLD was matching COLD and COLD_COOL.  Only want exact match.

## Changes in Version 1.1.0.9007 (2018-06-12)

* Update Rules.xlsx.
    + Level 3 (Hi and Lo), pi_BCG_att56 changed to pt_BCG_att1i23.
    + Change involved both direction and thresholds.

## Changes in Version 1.1.0.9006 (2018-06-12)

* Update metric.values.
    + Add "cols2keep" to allow for extra fields in output.
    + Need extra fields for qc.checks
* Update qc.checks.
    + Update example with cols2keep from metric.values example.

## Changes in Version 1.1.0.9005 (2018-06-11)

* Update bio data to be used for the model; Issue #7
    + Collapse mites to Order.

## Changes in Version 1.1.0.9004 (2018-06-08)

* Update extdata/Rules.xlsx; Issue #7

## Changes in Version 1.1.0.9003 (2018-06-08)

* metric.values dd nt, pi, and pt metrics; Issue #7
    + NonIns_BCG_attr456
    + NonInsJugaRiss_BCG_attr456

## Changes in Version 1.1.0.9002 (2018-06-08)

* QC Flags, Add pi_dom02.  Issue #7
* Update format of NEWS.
* metric.values, row_number to dplyr::row_number for dominant metrics.

## Changes in Version 1.1.0.9001 (2018-06-08)

* metric.values, Issue #7
    + Update metrics from NonClump to NoJugaRiss.


## Changes in Version 1.1.0.0000 (2018-03-19)

* Release new and fully QCed version.

## Changes in Version 1.0.0.9002 (2018-03-19)

* Metric.values
    + Fix metrics with *not equal* functions.
    + Ensure use is.na(x)==TRUE rather than x==NA
    + When have more than one condition for a data column need to be in the same statement.
    + Better define clumpy taxa for 2 metrics.
* BCG.Level.Membership
    + Fixed combination of rules by level.  Missed a column and was getting duplicates for alt rules that changed the outcome.
* BCG.Level.Assignment
    + Replace For loop with apply.
    + Add special conditions for ties (2 levels with 0.5) and primary level is 1 (no 2nd level).

## Changes in Version 1.0.0.9001 (2018-03-18)

* Metric.values
    + Fix non-clumpy metrics.
    + Remove extra data frames for dominance metrics.  Slight speed improvement.  Issue #5.
* Fix version numbers in NEWS.
* Updated ReadMe with only the relevant packages.  Issue #4.

## Changes in Version 1.0.0.0000 (2018-03-18)

* Only metric value issues left to resolve with test data set.  Rest of the calculations are ok.

## Changes in Version 0.1.0.9019 (2018-03-18)

* Rules.xlsx
    + Fill in missing cells.
* metric.values.R
    + Update FFG abbreviations to match example data.
    + Dominant N and special Dominant N metrics.
    + Define pipe as referncing dplyr.  Slows down the functions but keeps dplyr from needing to be loaded.
* SiteType values to lowercase.
    + metric.values.R
    + BCG.Metric.Membership.R
    + BCG.Level.Membership.R
    + BCG.Level.Assignment.R
* Consistent terminology; don't mix ".", " ", and "_".  Use "_".
    + METRIC_NAMES
    + METRIC_VALUES
    + SITE_TYPE (and change from REGION)
    + RULE_TYPE
    + NUMERIC_RULES
    + LIFE_CYCLE
    + LONG_LIVED
    + THERMAL_INDICATOR
    + NAME_WIDE
* Rebuild PacNW master taxa list with changes.
    + Consistent names (columns and categories).
* Example PacNW data.
    + Consistent names (columns).
* QC Checks
* Update Vignette.
* All functions and examples seem to be generating the proper outputs.  When QC only a few metrics are off.  But enough that no final results are correct (that is, all levels, site types, and samples are affected but the remaning QC issues).  Non-Insect and modified dominant 2 metrics are the 4 that have issues.

## Changes in Version 0.1.0.9018 (2018-03-17)

* BCG.Level.Membership.R
    + Correct function to consider preceeding levels when assigning membership scores.

## Changes in Version 0.1.0.9017 (2018-03-16)

* metric.values.R
    + Added Thermal Indicator (TI) metrics.
* Updates to PacNW bug data and rules files (xlsx).
    
## Changes in Version 0.1.0.9016 (2018-03-16)

* metric.values.R
    + Update pt_BCG_att1i2 and 1i23 metrics (numerator not capturing "i" taxa).
* Update Rules.xlsx
    
## Changes in Version 0.1.0.9015 (2018-03-16)

* Vignette didn't update in previous version.

## Changes in Version 0.1.0.9014 (2018-03-15)

* Tweak "data.R".  Not complete!
    + Added example save code for master taxa list.
* Update BCG.Metric.Membership.
    + Add some error checking.
    + Colnames to upper case.
* Updated Rules.xlsx
    + Percent metrics convered to 0-1 values to match metric.values calculations.
    + Fixed Index Name for some records.
* metric.values.R
    + Updated examples, details, and description.
* Update BCGcalc Vignette.
* Update Readme.
* Update BCG.Level.Membership

## Changes in Version 0.1.0.9013 (2018-03-14)

* Added PacNW benthic master taxa list to data.
* Created "data.R".  Not complete!

## Changes in Version 0.1.0.9012 (2018-03-08)

* Update PacNW example data to use TRUE/FALSE for Excluded and NonTarget fields.
* Updated metric.values
    + Moved common data munging to earlier in function so applies to all communities.
    + Enabled NonTarget exclusion.  NonTarget column now required.

## Changes in Version 0.1.0.9011 (2018-03-08)

* Completed lingering edits for "Tier" to "Level".
* Completed BCG.Level.Membership function.

## Changes in Version 0.1.0.9010 (2018-03-07)

* Remove "Tier" wording in package and rename some functions to better reflect usage.
* Include example code for saving results for each function.
* Change Region to SiteType in code and example data files.

## Changes in Version 0.1.0.9009 (2018-03-05)

* Dom02 special metric placeholder.
* Update Rule table for Level assignments.
* BCG.Levels function.
* Fixed error in sequencing in BCG.Membership function.

## Changes in Version 0.1.0.9008 (2018-03-05)

* Add last of the PacNW specific metrics to metric.values.
* Added Vignette.

## Changes in Version 0.1.0.9007 (2018-03-05)

* Check clean up.
* Convert data imports from development folder to pacakge folder with system.file.
* Add data for rarify function.

## Changes in Version 0.1.0.9006 (2018-03-04)

* BCG.Tiers function.

## Changes in Version 0.1.0.9005 (2018-03-04)

* Add BCG.Membership function.  Return is in long format.
    + Not QCed but has some known issues.

## Changes in Version 0.1.0.9004 (2018-03-03)

* Add subtitles, date, author to YAML header to NEWS and README.
* Add "flags" Excel file to raw-data for QC checks.  Used in qc.checks function (new).
* Added 4 new ni metrics to metric values to be used for QC checks (flags).
* qc.checks.R (incomplete but works).
* Add BCGcalc.R to describe the package.
* Update documentation headers in R files.

## Changes in Version 0.1.0.9003 (2018-02-20)

* metric.values.R, benthos
    + Added BCG attribute metrics.
    + Added outlining for readability in code.
    + More metrics and clean up of file.
    + Prepare example data for metric calculation.
    + Added na.rm=TRUE to n\_distinct and sum in summarise.

##  Changes in Version 0.1.0.9002 (2018-02-16)

* Add parts without any modification from `MBSStools` package.
    + rarify function
    + metric.values function
    + included data folder and files
    + added Excel file to extdata with metric naming explanation

## Changes in Version 0.1.0.9001 (2018-02-16)

* Updated files for package.
    + Readme, NEWS, DESCRIPTION
    + Not complete.

## Changes in Version 0.1.0 (2018-02-16)

* Created GitHub repository

