---
title: "BCGcalc-NEWS"
author: "Erik.Leppo@tetratech.com"
date: "`r Sys.time()`"
output: github_document
  #md_document:
   # variant: markdown_github
---

<!-- NEWS.md is generated from NEWS.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "NEWS-"
)
```

```{r, echo = FALSE}
cat(paste0("Last Update: ",Sys.time()))
```

# Planned Features

* Attribute calculation

* Tier membership

* Model calculation

* More/special metrics.

* Add excluded taxa flag to samples based on levels of taxonomy.  Will not be completely accurate.  Will have many false positives and false negatives.  That is, if there are only organisms at the higher level in a sample it will fail.  And if there are no organisms of the target group it will pass.
    + Pass, Fail on each. And overall flag
    + Chiro (tribe and family)
    + Oligo
    + trombidiformes

* Hi/Lo models.
    + parameter to run "correct" model or all models for each site.
    + Have a field for site type.  So can override gradient.
    
* Report - QC, use DataExplorer.

* Table of metrics for each model with DataExplorer report.

# Version History

## v0.1.0.9007
2018-03-05

* Check clean up.
* Convert data imports from development folder to pacakge folder with system.file.
* Add data for rarify function.

## v0.1.0.9006
2018-03-04

* BCG.Tiers function.

## v0.1.0.9005
2018-03-04

* Add BCG.Membership function.  Return is in long format.
    + Not QCed but has some known issues.

## v0.1.0.9004
2018-03-03

* Add subtitles, date, author to YAML header to NEWS and README.
* Add "flags" Excel file to raw-data for QC checks.  Used in qc.checks function (new).
* Added 4 new ni metrics to metric values to be used for QC checks (flags).
* qc.checks.R (incomplete but works).
* Add BCGcalc.R to describe the package.
* Update documentation headers in R files.

## v0.1.0.9003
2018-02-20

* metric.values.R, benthos
    + Added BCG attribute metrics.
    + Added outlining for readability in code.
    + More metrics and clean up of file.
    + Prepare example data for metric calculation.
    + Added na.rm=TRUE to n\_distinct and sum in summarise.

## v0.1.0.9002
2018-02-16

* Add parts without any modification from `MBSStools` package.
    + rarify function
    + metric.values function
    + included data folder and files
    + added Excel file to extdata with metric naming explanation

## v0.1.0.9001
2018-02-16

* Updated files for package.
    + Readme, NEWS, DESCRIPTION
    + Not complete.

## v0.1.0
2018-02-16

* Created GitHub repository

