---
title: "Vignette, BCGcalc"
author: "Erik.Leppo@tetratech.com"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette, QW}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
<!-- Data is in vignettes\data folder  -->
```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```
# Purpose
The `BCGcalc` package was created to enable users to implement a model created Biological Condition Gradient and apply it to their data.

This vignette will cover the basics going from raw data to model results.

# Tasks
1. Metric calculation

2. Metric scoring (membership)

3. Level membership

4. Level assignment

5. Other

## Metric Calculation
The `BCGcalc` package includes some example that will be used in this example.

THe funtion metric.values generates 80+ different metrics for multi-metric indices as well as BCG specific metrics.  All metrics are calculated and reported back to the user in a data frame.

```{r MetricValues_Calc, eval=FALSE}
# Metrics, BCG, Bugs
library(BCGcalc)
library(readxl)

# PACIFIC NW
df.samps.bugs <- read_excel(system.file("./extdata/Data_BCG_PacNW.xlsx"
                                       , package="BCGcalc"))
myDF <- df.samps.bugs

# calculate
df.metric.values.bugs <- metric.values(myDF, "bugs")
View(df.metric.values.bugs)

```

Other tasks can be performed on the data once it is returned as a data frame.

```{r MetricValues_Extra, eval=FALSE}
library(BCGcalc)
library(reshape2)
library(DataExplorer)

# Convert to long format
df.long <- reshape2::melt(df.metric.values.bugs, id.vars=c("SAMPLEID", "INDEX_NAME", "REGION")
                          , variable.name="metric.name", value.name="metric.value")
# Export for QC
write.table(df.long, "metric.values.tsv", col.names=TRUE, row.names=FALSE, sep="\t")

# DataExplorer Report
create_report(df.metric.values.bugs, "DataExplorer_Report_MetricValues.html")
create_report(df.samps.bugs, "DataExplorer_Report_BugSamples.html")
```

## Metric Scoring (Membership)
Metrics are scored according to a user defined table.  The membership scoring is BCG specific and results in a score in the range of 0 to 1.

The results are returned in a data frame in the long format.

```{r Import, eval=FALSE, echo=TRUE}
library(BCGcalc)
library(readxl)
library(knitr)

# Calculate Metrics
df.samps.bugs <- read_excel(system.file("./extdata/Data_BCG_PacNW.xlsx"
                                        , package="BCGcalc"))
myDF <- df.samps.bugs
df.metric.values.bugs <- metric.values(myDF, "bugs")

# Import Rules
df.rules <- read_excel(system.file("./extdata/Rules.xlsx"
                             , package="BCGcalc"), sheet="BCG_PacNW_2018") 

# Run function
df.Metric.Membership <- BCG.Metric.Membership(df.metric.values.bugs, df.rules.PacNW)

# show results
#View(df.Metric.Membership)

tbl.rules <- df.rules
tbl.rules.caption <- "Membership rules."

kable(tbl.rules, caption=tbl.rules.caption)

```

## Level Membership
The levels are determined according to the rules for a specific BCG model (e.g., Pacific Northwest low gradient).  The rules assign values (0-1) based on metric values.

```{r LevelMembership, eval=FALSE}
library(readxl)

# Calculate Metrics
df.samps.bugs <- read_excel(system.file("./extdata/Data_BCG_PacNW.xlsx"
                                        , package="BCGcalc"))
myDF <- df.samps.bugs
df.metric.values.bugs <- metric.values(myDF, "bugs")

# Import Rules
df.rules <- read_excel(system.file("./extdata/Rules.xlsx"
                             , package="BCGcalc"), sheet="BCG_PacNW_2018") 

# Calculate Metric Memberships
df.Metric.Membership <- BCG.Metric.Membership(df.metric.values.bugs, df.rules)

# Calculate Level Memberships
df.Level.Membership <- BCG.Level.Membership(df.Metric.Membership, df.rules)

# Show results
View(df.Level.Membership)
```

## Level Assignment
The level assignments are done according the values of the rules and metric level memberships.  The maximum level is designated as the primary level and the next highest value is assigned as the secondary level.

There is some QC to ensure the total membership equals 1 and a field to note if the assigned nears are a tie or close.
```{r LevelAssignment, eval=TRUE, echo=TRUE}
library(BCGcalc)
library(knitr)

# construct a dummy dataset
L1 <- rep(0, 10)
L2 <- c(0.4, 0, 0.4, rep(0,7))
L3 <- c(0.6, 0, 0.6, 0, 0.42, 0, 1, 1, 0.22, 0.33)
L4 <- c(0, 0.9, 0, 0, 0.58, 0.05, 0, 0, 0.78, 0.67)
L5 <- c(0, 0.1, 0, 1, 0, 0.95, rep(0,4))
L6 <- rep(0, 10)
SampleID <- LETTERS[1:10]
df.Level.Membership <- as.data.frame(SampleID, stringsAsFactors=FALSE)
df.Level.Membership[,"L1"] <- L1
df.Level.Membership[,"L2"] <- L2
df.Level.Membership[,"L3"] <- L3
df.Level.Membership[,"L4"] <- L4
df.Level.Membership[,"L5"] <- L5
df.Level.Membership[,"L6"] <- L6

# Run Function
df.Levels <- BCG.Level.Assignment(df.Level.Membership)

# # Show Results
# View(df.Levels)
# 
# # Save Results
# write.table(df.Levels, "Levels.tsv"
#             , row.names=FALSE, col.names=TRUE, sep="\t")

kable(df.Levels, caption="BCG Tier Assignments.")
```


## Other Tasks
There are some ancillary tasks that can be completed with the data that are a necessary part of data analysis and are included in the `BCGcalc` package for convenience.

### Master Taxa List
The Pacific Northwest benthic macroinvertebrate master taxa list is included for reference and can be saved to a file.

```{r BMT, eval=FALSE}
library(BCGcalc)

write.csv(TaxaMaster_Ben_BCG_PacNW, "TaxaMaster_Ben_BCG_PacNW_20180314.csv")
```

### Subsample (rarify)
Subsample (rarify) a biological sample to a fixed count.

Takes as an input a 3 column data frame (SampleID, TaxonID, Count) and returns a similar dataframe with revised Counts.

The other inputs are subsample size (target number of organisms in each sample) and seed. The seed is given so the results can be reproduced from the same input file. If no seed is given a random seed is used.  An easy seed to remember is the date of admission to the Union for each state (e.g., Washinton is 18891111).  These values can be found on Wikipedia on the right sidebar.

Returns a data frame with the same three columns but the abund field has been modified so the total count for each sample is no longer above the target (subsiz).

Code from USEPA Corvallis John Van Sickle R code for RIVPACS (v1.0, 2005-06-10).  Tweaked for addition of seed.

```{r Other_Rarify, eval=TRUE, echo=TRUE}
library(BCGcalc)
library(knitr)

# load bio data
DF.biodata <- data_bio2rarify
#dim(DF.biodata)
#View(DF.biodata)

# subsample
mySize <- 200
Seed.OR <- 18590214
Seed.WA <- 18891111
bugs.mysize <- rarify(inbug=DF.biodata, sample.ID="SampRep"
                     ,abund="Count",subsiz=mySize, mySeed=Seed.WA)
#dim(bugs.mysize)
#View(bugs.mysize)

# Compare
df.compare <- merge(DF.biodata, bugs.mysize, by=c("SampRep", "tax"))
#View(df.compare)
 
# save the data
#write.table(bugs.mysize,paste("bugs",mySize,"txt",sep="."),sep="\t")

tbl.compare <- head(df.compare)
tbl.compare.caption <- "First few rows of original and rarified data."

kable(tbl.compare, caption=tbl.compare.caption)

```

### QC Checks
Specific to some programs there are QC checks on the samples that should be performed.  For example, minimum and maximum number of organisms to be a valid sample.

This function uses an imported data frame to compare calcualted metrics and gives a Pass or Fail evaluation.

The QC checks are matched against the data on Index Name, Region, and Metric Name.  Only those that match are evaluated.  Valid symbols are those used by R; <, >, <=, >=, ==, and !=.
```{r Other_QCCheck, eval=TRUE, echo=TRUE}
library(readxl)
library(BCGcalc)
library(knitr)

# Calculate Metrics
df.samps.bugs <- read_excel(system.file("./extdata/Data_BCG_PacNW.xlsx"
                                        , package="BCGcalc"))
myDF <- df.samps.bugs
df.metric.values.bugs <- metric.values(myDF, "bugs")

# Import Checks
df.checks.PacNW <- read_excel(system.file("./extdata/MetricFlags.xlsx"
                                          , package="BCGcalc"), sheet="Flags") 

# Run function
df.flags <- qc.checks(df.metric.values.bugs, df.checks.PacNW)

# Show QC checks table
tbl.checks <- df.checks.PacNW
tbl.checks.caption <- "QC Checks"

kable(tbl.checks, caption = tbl.checks.caption)

# Show results summary
tbl.results <- table(df.flags[,"CheckName"], df.flags[,"Flag"])
tbl.results.caption <- "QC Check Results"

kable(tbl.results, caption = tbl.results.caption)
```


